---
title: "DP"
author: "Will Townes"
date: "March 19, 2016"
output: html_document
---

```{r}
library(mvtnorm)
library(ggplot2)
```

First we generate some fake data from a mixture model of three two-dimensional Gaussians. In the first simulation, we assume isotropic, known variance for all clusters.

```{r}
genCluster<-function(n,id,mu=c(0,0),sigma=diag(2)){
  d<-data.frame(rmvnorm(n,mean=mu,sigma=sigma))
  d$id<-id
  return(d)
}
red<-genCluster(20,"1") #sigma=cbind(c(2,-.5),c(-.5,2))
blue<-genCluster(20,"2",c(2,2))
green<-genCluster(20,"3",c(-2,-2))
dat<-rbind(red,blue,green)
ggplot(data=dat)+geom_point(aes(x=X1,y=X2,color=id),size=2)
```

Now we create a Gibbs Sampler for the Dirichlet Process. We specify a gaussian likelihood with known isotropic variance and unknown mean parameter. 

The prior on the mean parameters is that they are iid from a dirichlet process with (variously specified) concentration parameters and base measure gaussian with mean zero and variance 10.

```{r}

```
