---
title: "LDS_Harmonic.Rmd"
author: "Will Townes"
date: "May 10, 2016"
output: html_document
---

This notebook picks up where "LDS.Rmd" leaves off.

```{r}
library(ggplot2)
library(pander)
source("../code/func/sampler/lds/lds_util.R")
source("util.R")
```


### Harmonic Motion with Known Dynamical Parameters

Both the linear and parabolic trajectories have trends. A stationary linear dynamical system is exemplified by harmonic motion (spring motion). To facilitate visualization, we assume a one dimensional system taking positions $q_t=q(t)$ defined by the following differential equation:
$$q''(t) = -k \left(q(t)-L\right)$$
Where $k$ is a constant related to material properties of the spring and $L$ is the equilibrium position. This is a one dimensional, second order system for position, and can be represented as a two-dimensional system of first order differential equations.
Converting to the discretized state space model, let $x_t=(q_t,v_t)$ be defined as before. We have
$x_t\sim\mathcal{N}(A x_{t-1}+B,\Sigma)$
where
$$A=\begin{pmatrix}1&\Delta t\\ -k\Delta t & 1\end{pmatrix}$$
To account for acceleration, $B=(0,k L \Delta t )'$ . We now simulate various patterns of harmonic motion to illustrate the effect of the parameters.

```{r}
k<-.1; L<-100
Tmax<-100
z<-rep(1,Tmax)
A<-diag(2)
A[1,2]<- 1
A[2,1]<- -k
theta<-list(list(A=A,B=c(0,k*L),Sigma=5*diag(2)))
lambda<-list(C=matrix(c(1,0),nrow=1),R=matrix(100000,nrow=1,ncol=1))
xy<-gen_lds(z,c(L-200,0),theta,lambda)
x<-xy[["x"]]
y<-xy[["y"]]
time<-1:Tmax
plot(time,x[,1],type="l",main="Noisy, Undamped Harmonic Motion",xlab="Time",ylab="Position")
points(time,y[,1])
```

This kind of "blowing up" behavior is an artifact of the discretization of the differential equation. It can be improved by using damped harmonic motion:
$$u''(t) = -k u(t) - \delta u'(t)$$
The $\delta$ parameter causes the acceleration to tend toward zero. In terms of the linear dynamical system representation, this corresponds with replacing the one on the diagonal with $1-\delta$ in the fourth position (the vertical velocity position).

```{r}
make_harmonic_matrices<-function(k,L,delta,Sigma_var=2){
  # return the A, B, Sigma matrices for harmonic motion
  A<-matrix(c(1,-k,1,1-delta),nrow=2)
  #A[1,2]<- 1
  #A[2,1]<- -k
  #A[2,2]<-1-delta
  return(list(A=A,B=c(0,k*L),Sigma=Sigma_var*diag(2)))
}
#k<-.01; L<-0; delta<-.1
Tmax<-500
z<-rep(1,Tmax)
theta<-list(make_harmonic_matrices(.01,0,.1))
lambda<-list(C=matrix(c(1,0),nrow=1),R=matrix(10,nrow=1,ncol=1))
xy<-gen_lds(z,c(L+10,0),theta,lambda)
x<-xy[["x"]]
y<-xy[["y"]]
time<-1:Tmax
plot(time,x[,1],type="l",main="Noisy, Damped Harmonic Motion",xlab="Time",ylab="Position")
points(time,y[,1])
```

### Switching Harmonic Motion

We now consider a scenario where there are three kinds of harmonic motion, the first one with $k=.01, L=0, \delta=.1$, the second with $k=.01, L=100, \delta=.2$, and the third with $k=.1, L=0, \delta=.1$.

```{r}
Tmax<-100
z<-rep(c(1,2,3),each=Tmax)
#modes<-factor(c(1,z))
Sigma_var<-1
theta[[1]]<-make_harmonic_matrices(.01,30,.5,Sigma_var)
theta[[2]]<-make_harmonic_matrices(.01,-10,.5,Sigma_var)
theta[[3]]<-make_harmonic_matrices(.1,5,.2,Sigma_var)
xy<-gen_lds(z,c(0,0),theta,lambda)
x<-xy[["x"]]
y<-xy[["y"]]
Tmax<-Tmax*3
time<-1:Tmax
plot(time,x[,1],type="l",main="Noisy, Damped Harmonic Motion",xlab="Time",ylab="Position")
points(time,y[,1])
colnames(x)<-c("position","velocity")
colnames(y)<-c("position")
xs<-rLDS(100,y,z,theta,lambda,x0_return=TRUE) #100 samples from the joint posterior
xs<-lapply(xs,function(x){cbind(x,c(0,time))})
xs<-rLDS_melt(xs,c(colnames(x),"time"))
x_true<-as.data.frame(x)
#x_true$time<-time
y_obs<-as.data.frame(y)
ggplot(data=xs,aes(x=time,y=position))+geom_path(aes(group=id),colour="grey")+geom_path(data=x_true,size=0.8,aes(colour=factor(z)))+geom_point(data=y_obs,size=1.1)+theme_bw()+xlab("Time")+ylab("Position")+ggtitle("Harmonic Motion with Known Dynamical Parameters")#+scale_fill_hue(c=85,l=85)
```

### Harmonic Motion with Unknown Dynamical Parameters

```{r}
theta_true<-theta
K<-3 #number of dynamical modes
D<-2 #latent dimension
#x_0<-x0_init(x)[1:D] #cheating here and using the real values, must be a better way?
#initialize from default settings based on the data.
hp<-hyper_init(y,D=D) 
#alternatively, adjust priors by hand to force stable dynamics
#hp<-list(kappa0=10,S0_df=10,R0_df=10,K_A=1*diag(D))
#hp<-hyper_init(y,D=D,hp)
#hp$S0<-diag(diag(hp$S0)) #force diagonal
#hp$R0<-diag(diag(hp$R0))
theta_hat<-replicate(K,theta_init(hp),simplify=FALSE)
lambda_hat<-lambda_init(hp)
lambda_hat[["R"]]<-lambda[["R"]] #cheat and use the true value for R

res<-sample_lds_posterior(y,z,theta_hat,lambda_hat,hp)
xhat<-res[["x"]]
theta_hat<-res[["theta"]]
xs<-lapply(xhat,function(x){cbind(x,c(0,time))})
xs<-rLDS_melt(xs,c(colnames(x),"time"))
#visualize sample trajectory
ggplot(data=xs,aes(x=time,y=position))+geom_path(aes(group=id),colour="grey")+geom_path(data=x_true,size=0.8,aes(colour=factor(z)))+geom_point(data=y_obs,size=1.1)+theme_bw()+xlab("Time")+ylab("Position")+ggtitle("Harmonic Motion with Unknown Dynamical Parameters")#+scale_fill_hue(c=85,l=85)
```

```{r}
prior_mean<-list()
prior_mean<-list(A=hp$M_A,B=hp$M_B,Sigma=hp$S0/(hp$S0_df-D-1))
mse_table<-data.frame(mode=rep(1:K,each=3),param=rep(c("A","B","Sigma"),K),MSE_true=NA,MSE_prior=NA)
for(zval in 1:K){
  for(parval in c("A","B","Sigma")){
    par_post_mean<-list_mean(theta_hat,function(x){x[[zval]][[parval]]})
    truval<-theta_true[[zval]][[parval]]
    mseval<-mse(par_post_mean,truval)
    mseprior<-mse(par_post_mean,prior_mean[[parval]])
    mse_table[mse_table$mode==zval & mse_table$param==parval, c("MSE_true", "MSE_prior")] <- c(mseval,mseprior)
  }
}
mse_table[,c("MSE_true","MSE_prior")]<-signif(mse_table[,c("MSE_true","MSE_prior")],3)
pander(mse_table)
```
```